---
title: "Assignment 1 - Language Development in ASD - part 3"
author: "Riccardo Fusaroli"
date: "August 10, 2018"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

locpath= "C:/Users/nanna/OneDrive - Aarhus universitet/3. Semester/Experimental Methods 3/LanguageDevelopmentASD"

setwd(locpath)
getwd()


library(pacman)
p_load(lmerTest, pastecs, ggplot2, tidyverse, gdata, MuMIn, effects, stringr)


```

## Welcome to the third exciting part of the Language Development in ASD exercise

In this exercise we will delve more in depth with different practices of model comparison and model selection, by first evaluating your models from last time, then learning how to cross-validate models and finally how to systematically compare models.

N.B. There are several datasets for this exercise, so pay attention to which one you are using!

1. The (training) dataset from last time (the awesome one you produced :-) ).

2. The (test) datasets on which you can test the models from last time:

* Demographic and clinical data: https://www.dropbox.com/s/ra99bdvm6fzay3g/demo_test.csv?dl=1
* Utterance Length data: https://www.dropbox.com/s/uxtqqzl18nwxowq/LU_test.csv?dl=1
* Word data: https://www.dropbox.com/s/1ces4hv8kh0stov/token_test.csv?dl=1

```{r, include=FALSE}

#Loading the new data 

demo_test <- read.csv("demo_test.csv")
  
token_test <- read.csv("token_test.csv")

lu_test <- read.csv("lu_test.csv")

```



```{r, include=FALSE}
#Renaming the data sets to have the same headline for the subject = ID

demo_test <- rename.vars(demo_test, "Child.ID", "ID", info=T)
lu_test <- rename.vars(lu_test, "SUBJ", "ID", info=T)
token_test <- rename.vars(token_test, "SUBJ", "ID", info=T)


#Renaming the visit to have the same 
lu_test <- rename.vars(lu_test, "VISIT", "Visit", info=T)
token_test <- rename.vars(token_test, "VISIT", "Visit", info=T)

#Homogeneize the way visit is reported, use of stringr to extract only numbers
lu_test$Visit <- str_extract(lu_test$Visit, "[1-6]")
token_test$Visit <- str_extract(lu_test$Visit, "[1-6]")


#Removing all punctuations in the ID column 
demo_test$ID <- str_replace_all(demo_test$ID, "[:punct:]", "")
lu_test$ID <- str_replace_all(lu_test$ID, "[:punct:]", "")
token_test$ID <- str_replace_all(token_test$ID, "[:punct:]", "")

#Selecting the variables needed
demo_test_sub <- select(demo_test, c(ID, Visit, Ethnicity, Diagnosis, Gender, Age, ADOS,  MullenRaw, ExpressiveLangRaw))

lu_test_sub <- select(lu_test, c(ID, Visit, MOT_MLU, MOT_LUstd, CHI_MLU, CHI_LUstd))

token_test_sub <- select(token_test, c(ID, Visit, types_MOT, types_CHI, tokens_MOT, tokens_CHI))


library(plyr)
#Using join to merge to three datasets by ID and Visit
test <- join(demo_test_sub, lu_test_sub)
test <- join(test, token_test_sub)


#Ados for only visit 1
#Using subset to make a dataset only consisting of the data from visit 1
subset_test <- subset(test, Visit == 1, select = c(ID, ADOS, MullenRaw, ExpressiveLangRaw))

#Changing the names in the new data set 
subset_test <- rename.vars(subset_test, c("ADOS", "MullenRaw", "ExpressiveLangRaw"), c("ADOS1", "NonverbalIQ", "VerbalIQ"))

#Merging the two datasets using join 
test <- join(test, subset_test)

#Using select to make a clean dataset with variables in the right order
clean_test <- select(test, c(ID, Diagnosis, Visit, Gender, Age, Ethnicity,  ADOS1, NonverbalIQ, VerbalIQ, CHI_MLU, MOT_MLU, CHI_LUstd, MOT_LUstd, types_CHI, types_MOT,  tokens_CHI, tokens_MOT))

#Anonymous kids
clean_test$ID <- as.factor(clean_test$ID)
clean_test$ID <- as.numeric(clean_test$ID)

#Turning the 1s and 2s into M and F in the gender variable
clean_test$Gender <- ifelse(clean_test$Gender == "1", "M", "F")

#Turning the As and Bs into ADS and TD in the diagnosis variable using ifelse (short way)
clean_test$Diagnosis <- ifelse(clean_test$Diagnosis == "A", "ASD", "TD")

#Making a csv. file named LanguageASD.csv to WD
write.csv(clean_test, file = "LanguageASD_test.csv")

```

#######Start loading from here#######



Loading CSV's

```{r, include=FALSE}
data = read.csv("LanguageASD.csv")
data_test = read.csv("LanguageASD_test.csv")
```

Changing names in the traning data
```{r, include=FALSE}
data <- rename.vars(data, "Diagnosis", "diagnosis")
data <- rename.vars(data, "Gender", "gender")
data <- rename.vars(data, "Age", "age")
data <- rename.vars(data, "Ethnicity", "ethnicity")
data <- rename.vars(data, "Visit", "visit")
data <- rename.vars(data, "ADOS1", "ados1")
data <- rename.vars(data, "MullenRaw1", "nonverbalIQ")
data <- rename.vars(data, "ExpressiveLangRaw1", "verbalIQ")

data <- data[-c (1, 68, 69, 130, 131, 132, 276, 277, 284, 285), ]

```

Changing names in the test data
```{r, include=FALSE}
data_test <- rename.vars(data_test, "Diagnosis", "diagnosis")
data_test <- rename.vars(data_test, "Gender", "gender")
data_test <- rename.vars(data_test, "Age", "age")
data_test <- rename.vars(data_test, "Ethnicity", "ethnicity")
data_test <- rename.vars(data_test, "Visit", "visit")
data_test <- rename.vars(data_test, "ADOS1", "ados1")
data_test <- rename.vars(data_test, "NonverbalIQ", "nonverbalIQ")
data_test <- rename.vars(data_test, "VerbalIQ", "verbalIQ")

```


### Exercise 1) Testing model performance

How did your models from last time perform? 

In this exercise you have to compare the results on the training data () and on the test data. Report both of them. 
Compare them. 

Discuss why they are different.


- recreate the models you chose last time (just write the model code again and apply it to your training data (from the first assignment))

- calculate performance of the model on the training data: root mean square error is a good measure. (Tip: google the function rmse())

- create the test dataset (apply the code from assignment 1 part 1 to clean up the 3 test datasets)

- test the performance of the models on the test data (Tips: google the functions "predict()")


- optional: predictions are never certain, can you identify the uncertainty of the predictions? (e.g. google predictinterval())


formatting tip: If you write code in this document and plan to hand it in, remember to put include=FALSE in the code chunks before handing in.

```{r, include=FALSE}
p_load(Metrics)


#Removing NAs in data sets
data <- na.omit(data, cols = "CHI_MLU")
data_test <- na.omit(data_test, cols = "CHI_MLU")


#The best model from the last assignment
model_train <- lmer(CHI_MLU ~ visit + I(visit^2) + I(visit^3)+ ados1 + MOT_MLU + verbalIQ + (1+visit + I(visit^2)+ I(visit^3) | ID), data = data, REML=FALSE)



#Root mean square error on how the model perform on the traning data
rmse(data$CHI_MLU, predict(model_train))
##[1] 0.2529832


#Very bad prediction on the test set 
rmse(data_test$CHI_MLU,predict(model_train, data_test))
##[1] 0.780272


```

Our cubic model from last assignment that we chose as the best model (pseudocode: mean length of utterance ~ visit + visit^2 + ados + mothers mlu + (visit + visit^2 + visit^3 | ID)) produces a root mean square error of 0.25. The root mean square error increases quite a lot when applying the model on the test set. It goes from 0.25 to 0.78.

Our model is basically bad at predicting, generalizing to the population. It is possibly a sign of overfitting to the initial train data set. The explanatory power is high but the predictive power is weak. 



### Exercise 2) Model Selection via Cross-validation (N.B: ChildMLU!)

One way to reduce bad surprises when testing a model on new data is to train the model via cross-validation. 

In this exercise you have to use cross-validation to calculate the predictive error of your models and use this predictive error to select the best possible model.


- Use cross-validation to compare your model from last week with the basic model 
(Child MLU as a function of Time and Diagnosis, and don't forget the random effects!)


- (Tips): google the function "createFolds";  loop through each fold, train both models on the other folds and test them on the fold)


Which model is better at predicting new data: 

1) The one you selected last week or
2) the one chosen via cross-validation this week?

- Test both of them on the test data.
- Report the results and comment on them.

``` {r, include=FALSE}
p_load(vtreat)

#Creating the basic model
basic <- lmer(CHI_MLU ~ visit + diagnosis + (1 + visit | ID), data = data, REML = F)

#Creating our model from last week
bestmodel <- lmer(CHI_MLU ~ visit + I(visit^2) + I(visit^3)+ ados1 + MOT_MLU + verbalIQ + (1+visit + I(visit^2)+ I(visit^3) | ID), data = data, REML=FALSE)

```


Cross Validation 
```{r, include=FALSE}
# Get the number of rows in data
nRows <- nrow(data)

# Implement the 5-fold cross-fold plan
splitPlan <- kWayCrossValidation(nRows, 5, NULL, NULL)

# Examine the split plan
#str(splitPlan)

# Run the 5-fold cross validation plan from splitPlan on basic model
k <- 5 # Number of folds

#Setting the variable for the predictive values from the CV
data$pred.cv.lm <- 0 

#Creating the loop
for(i in 1:k) {
  split <- splitPlan[[i]]
  model <- lmer(CHI_MLU ~ visit + diagnosis + (1 + visit | ID), data = data[split$train, ], REML = F)
  data$pred.cv.lm[split$app] <- predict(model, newdata = data[split$app, ])
}


#RMSE for basic model
rmse(data$CHI_MLU, predict(basic))
##[1] 0.3380725

#Rmse of the cross-validation predictions
rmse(data$pred.cv.lm, data$CHI_MLU)
##[1] 0.5058554



# Run the 5-fold cross validation plan from splitPlan on our best model
data$pred.cv.best <- 0 

for(i in 1:k) {
  split1 <- splitPlan[[i]]
  model2 <- lmer(CHI_MLU ~ visit + I(visit^2) + ados1 + MOT_MLU + (1+visit + I(visit^2) | ID), data = data[split1$train, ], REML=FALSE)
  data$pred.cv.best[split1$app] <- predict(model2, newdata = data[split1$app, ])
  
}


# Get the rmse of the full model's predictions
rmse(data$CHI_MLU, predict(bestmodel))
##[1]  0.2529832

# Get the rmse of the cross-validation predictions
rmse(data$pred.cv.best, data$CHI_MLU)
##[1] 0.4515496


#How does the model predict the test data in reality?
rmse(predict(basic, data_test), data_test$CHI_MLU)
#[1] 1.068573

rmse(predict(bestmodel, data_test), data_test$CHI_MLU)
#[1] 0.780272


```

Another way to do CV  - just to check it out

The basic model
```{r, include=FALSE}
p_load("cvTools")

#k <- 10 #the number of folds

#Crate the folds
folds <- cvFolds(NROW(data), K=k)



#Set up column in dataframe
data$cv.pred <- rep(0,nrow(data))


#The loop
for(i in 1:k){
  train <- data[folds$subsets[folds$which != i], ] #Set the training set
  validation <- data[folds$subsets[folds$which == i], ] #Set the validation set

  newlm <- lmer(CHI_MLU ~ visit + diagnosis + (1 + visit | ID), data = train, REML = F) 
  #Get your new linear model (just fit on the train data)
    newpred <- predict(newlm,newdata=validation) 
  #Get the predicitons for the validation set (from the model just fit on the train data)

  data[folds$subsets[folds$which == i], ]$cv.pred <- newpred #Put the hold out prediction in the data set for later use
}

#How does the basic model perform
rmse(data$cv.pred, data$CHI_MLU)


```

The "best" model from last assignment
```{r, include=FALSE}

#preparing a column
data$cv.pred_best <- rep(0,nrow(data))


for(i in 1:k){
  train <- data[folds$subsets[folds$which != i], ] #Set the training set
  validation <- data[folds$subsets[folds$which == i], ] #Set the validation set

  newlmer <- lmer(CHI_MLU ~ visit + I(visit^2) + ados1 + MOT_MLU + (1+visit + I(visit^2) | ID), data = train, REML=FALSE)
  #Get your new linear model (just fit on the train data)
  
  newpred_best <- predict(newlmer, newdata=validation) 
  #Get the predicitons for the validation set (from the model just fit on the train data)

  data[folds$subsets[folds$which == i], ]$cv.pred_best <- newpred_best 
  #Put the hold out prediction in the data set for later use
}


rmse(data$cv.pred_best, data$CHI_MLU)

```

The basic linear model (pseudocode: mean length of utterance ~ visit + diagnosis + (visit | ID)) produces a root mean squared error of only 0.34, when predicting on the training data set. However, when the model is trained with a 5-fold cross validation, the root mean squared error increases to 0.51. 

As already mentioned earlier the cubic model, that we chose to be the best model in the previous assignment (pseudocode: mean length of utterance ~ visit + visit^2 + visit^3 + verbalIQ + ados + mother's MLU (visit + visit^2 + visit^3| ID)) produces a mean squared error of 0.28, when applied to the traning data. As with the linear model, the error increases when the model is trained with a 5-fold cross validation - to 0.46, respectively. 

First of all, this means that the cubic model is still better at predicting the traning data compared to the linear model.

When the two models' cross validation predictions are applied to the actual test data they both turn out to be quite far off. The linear model produces a root mean squared error of 1.07. The cubic model produces a rmse of 0.79. Thus, the cubic model is still performing better on the test data than the linear, however what worse than predicted with the traning data (both with and without cross validation of the model).


##Best model 

- Now try to find the best possible predictive model of ChildMLU, that is, the one that produces the best cross-validated results.


- Bonus Question 1: What is the effect of changing the number of folds? Can you plot RMSE as a function of number of folds?

- Bonus Question 2: compare the cross-validated predictive error against the actual predictive error on the test data

```{r, include=FALSE}
# Implement the 5-fold cross-fold plan
splitPlan <- kWayCrossValidation(nRows, 5, NULL, NULL)

k2 <- 5
data$pred.cv.create <- 0 



#Model with gender

for(i in 1:k2) {
  split.create <- splitPlan[[i]]
  model.create <- lmer(CHI_MLU ~ visit + I(visit^2) + ados1 + MOT_MLU + verbalIQ + (1+visit + I(visit^2) | ID) + (1| gender), data = data[split.create$train, ], REML=FALSE)
  
  data$pred.cv.create[split.create$app] <- predict(model.create, newdata = data[split.create$app, ])

}

rmse(data$pred.cv.create, data$CHI_MLU)
rmse(predict(model.create, data_test), data_test$CHI_MLU)
#[1] 0.4313872
#[1] 0.7579469


data$pred.cv.create1 <- 0 

# Loop using Nanna's random model
for(i in 1:k) {
  split.create1 <- splitPlan[[i]]
  model.create1 <- lmer(CHI_MLU ~ visit + I(visit^2) + I(visit^3) + ados1 + verbalIQ + MOT_MLU + types_CHI + (1+visit + I(visit^2)+ I(visit^3)| ID), data = data[split.create1$train, ], REML=FALSE)
  data$pred.cv.create1[split.create1$app] <- predict(model.create1, newdata = data[split.create1$app, ])
}


rmse(data$CHI_MLU, data$pred.cv.create1)
rmse(predict(model.create1, data_test), data_test$CHI_MLU)
#[1] 0.3881532
#[1] 0.5186018

data$pred.cv.create2 <- 0 


# Loop using the p-hacking model from ass. 2
for(i in 1:k) {
  split.create2 <- splitPlan[[i]]
  model.create2 <- lmer(CHI_MLU ~ visit + I(visit^2) + I(visit^3) + ID + nonverbalIQ + verbalIQ + types_CHI + MOT_MLU + (1+visit + I(visit^2)+ I(visit^3)| ID), data = data[split.create2$train, ], REML=FALSE)
  data$pred.cv.create2[split.create2$app] <- predict(model.create2, newdata = data[split.create2$app, ])
}

rmse(data$CHI_MLU, data$pred.cv.create2)
rmse(predict(model.create2, data_test), data_test$CHI_MLU)
#[1] 0.3877639
#[1] 0.5044019



```

Using Cross Validation to calculate accuracy of predictions and using the root mean squared error when applying the actual test data, the best predictive model is our p-hacked model from the previous assignment (pseudocode: Child MLU  ~ visit + I(visit^2) + I(visit^3) + ID + nonverbalIQ + verbalIQ + types_CHI + MOT_MLU + (1+visit + I(visit^2)+ I(visit^3)| ID). Interestingly, this model does not include ados or diagnosis. 


### Exercise 3) Assessing the single child

Let's get to business. This new kiddo - Bernie - has entered your clinic. This child has to be assessed according to his group's average and his expected development.

Bernie is one of the six kids in the test dataset, so make sure to extract that child alone for the following analysis.

```{r, include=FALSE}

#Find Bernie before the data was made anonymous
filter(test, ID == "Bernie")

#filter out Bernie form the data_test dataframe
bernie <- filter(data_test, ados1 == "7", verbalIQ == "28")


```


You want to evaluate:

- how does the child fare in ChildMLU compared to the average TD child at each visit? 

Define the distance in terms of absolute difference between this Child and the average TD.


(Tip: recreate the equation of the model: Y=Intercept+BetaX1+BetaX2, etc; input the average of the TD group for each parameter in the model as X1, X2, etc.).

```{r, include=FALSE}


#Creating the data needed to fill in the parameters for TD kids

#Filtering TD into its own data frama
TD_average <- data %>% 
  filter(diagnosis == "TD")

TD_average$visit <- as.numeric(TD_average$visit)

TD_visit <- TD_average %>% dplyr::group_by(visit) %>%
  dplyr::summarise(mean_TD = mean(CHI_MLU))

TD_ados <- TD_average %>%
  dplyr::group_by(visit) %>%
  dplyr::summarise(mean_TD = mean(ados1))

TD_motmlu <- TD_average %>%
  dplyr::group_by(visit) %>%
  dplyr::summarise(mean_TD = mean(MOT_MLU, na.rm =T))

TD_verbaliq <- TD_average %>%
  dplyr::group_by(visit) %>%
  dplyr::summarise(mean_TD = mean(verbalIQ, na.rm =T))

#Summary of bestmodel
summary(bestmodel)


intercept = -0.073966 
visit =  0.450687
visit2 = -0.035911
visit3 = -0.005445
ados1 =  -0.010203
MOT_MLU = 0.286858
verbalIQ = 0.060751


#Calculate TD ChildMlu using the model 
TD1 <- intercept + visit*TD_visit$visit[1] + visit2*(TD_visit$visit[1])^2 +visit3*(TD_visit$visit[1])^3 + ados1*TD_ados$mean_TD[1] + MOT_MLU*TD_motmlu$mean_TD[1] + verbalIQ*TD_verbaliq$mean_TD[1]

TD2 <- intercept + visit*TD_visit$visit[2] + visit2*(TD_visit$visit[2])^2 + +visit3*(TD_visit$visit[2])^3 + ados1*TD_ados$mean_TD[2] + MOT_MLU*TD_motmlu$mean_TD[2] + verbalIQ*TD_verbaliq$mean_TD[2]

TD3 <- intercept + visit*TD_visit$visit[3] + visit2*(TD_visit$visit[3])^2 + +visit3*(TD_visit$visit[3])^3 + ados1*TD_ados$mean_TD[3] + MOT_MLU*TD_motmlu$mean_TD[3] + verbalIQ*TD_verbaliq$mean_TD[3]

TD4 <- intercept + visit*TD_visit$visit[4] + visit2*(TD_visit$visit[4])^2 + +visit3*(TD_visit$visit[4])^3 + ados1*TD_ados$mean_TD[4] + MOT_MLU*TD_motmlu$mean_TD[4] + verbalIQ*TD_verbaliq$mean_TD[4]

TD5 <- intercept + visit*TD_visit$visit[5] + visit2*(TD_visit$visit[5])^2 + +visit3*(TD_visit$visit[5])^3 + ados1*TD_ados$mean_TD[5] + MOT_MLU*TD_motmlu$mean_TD[5] + verbalIQ*TD_verbaliq$mean_TD[5]

TD6 <- intercept + visit*TD_visit$visit[6] + visit2*(TD_visit$visit[6])^2 + +visit3*(TD_visit$visit[6])^3 + ados1*TD_ados$mean_TD[6] + MOT_MLU*TD_motmlu$mean_TD[6] + verbalIQ*TD_verbaliq$mean_TD[6]


#Turn into dataframe  
bernie$TD_MLU <- 0
bernie$TD_MLU <- c(TD1, TD2, TD3, TD4, TD5, TD6)

bernie$TD_MLU
bernie$CHI_MLU

bernie$TD_raw <- 0
bernie$TD_raw <- TD_visit$mean_TD

#Difference from Bernie and the models predictions of a TD
bernie$absolute_dif_model <- bernie$TD_MLU - bernie$CHI_MLU

#Difference from Bernie and the raw TD
bernie$absolute_dif_raw <- bernie$CHI_MLU - bernie$TD_raw



#Using the model to predict Bernie
bernie$predictions <- predict(bestmodel, bernie)

bernie_clean <- select(bernie, c(visit, CHI_MLU, predictions, TD_MLU, TD_raw, absolute_dif_model, absolute_dif_raw))
bernie_clean


```

- how does the child fare compared to the model predictions at Visit 6? Is the child below or above expectations? 

(tip: use the predict() function on Bernie's data only and compare the prediction with the actual performance of the child)

```{r, include=FALSE}


bernie$expectation <- (bernie$CHI_MLU - bernie$predictions)
bernie_clean$expectation <- bernie$expectation

##[1] -0.009466836 -0.679125251 -0.146376680 -0.605950913 -0.474317351  0.103381876

```

We have usen the raw data to calculate the averge of the kid with TD groupe by visit and compared the observations to Bernie, a ASD participant. 

When comparing Bernie to the average TD participant, Bernie is constantly outperforming the average TD. Take for instance visit 1, Bernie has a mean length of utterance on 1.98. The average TD has a MLU of 1.3.

At visit 6, the prediction of the model is a ChildMLU of 3.36, Bernie however, has a ChildMLU at 3.45. Bernie is better then the predicted values. The model is however fairly close at predciting correct.  When using the model to calculate the predictions on Bernie's data only, we see that Bernie's mean lenght of utterance is rather close to expectations at the first visit but in general below expectations at the first 5 visits. However, at visit 6 Bernie performed above expectations. 
This development can also be seen in the plot below, where the blue line indicated the predicitons obtained by the model and the red line is Bernie's actual performance. 


```{r}

bernie_clean

#Plotting Bernies predictions versus actualt performance 
theme_set(theme_bw())
ggplot(data = bernie, aes(x = visit)) + geom_smooth(aes(y=CHI_MLU), colour="red", method = "lm", formula = y ~ poly(x, 3), se = F) + 
         geom_smooth(aes(y=predictions), colour="blue", method = "lm", formula = y ~ poly(x, 3), se = F) +
  xlab('Number of Visit') +
  ylab("Bernie's Mean Lenght of Utterance")

```




### OPTIONAL: Exercise 4) Model Selection via Information Criteria
Another way to reduce the bad surprises when testing a model on new data is to pay close attention to the relative information criteria between the models you are comparing. Let's learn how to do that!

Re-create a selection of possible models explaining ChildMLU (the ones you tested for exercise 2, but now trained on the full dataset and not cross-validated).

Then try to find the best possible predictive model of ChildMLU, that is, the one that produces the lowest information criterion.

- Bonus question for the optional exercise: are information criteria correlated with cross-validated RMSE? That is, if you take AIC for Model 1, Model 2 and Model 3, do they co-vary with their cross-validated RMSE?

### OPTIONAL: Exercise 5): Using Lasso for model selection

Welcome to the last secret exercise. If you have already solved the previous exercises, and still there's not enough for you, you can expand your expertise by learning about penalizations. Check out this tutorial: http://machinelearningmastery.com/penalized-regression-in-r/ and make sure to google what penalization is, with a focus on L1 and L2-norms. Then try them on your data!

