---
title: "Assignment 1 - Language Development in ASD - part 4"
author: "Riccardo Fusaroli"
date: "August 10, 2017"
output: html_document
---

```{r + WD and packages }
locpath= "C:/Users/nanna/OneDrive - Aarhus universitet/3. Semester/Experimental Methods 3/LanguageDevelopmentASD"

setwd(locpath)
getwd()

library(pacman)
p_load(tidyverse, lmerTest, simr, gdata, lme4)

```


```{r + Load data }
data = read.csv("LanguageASD.csv")

data <- rename.vars(data, "Diagnosis", "diagnosis")
data <- rename.vars(data, "Gender", "gender")
data <- rename.vars(data, "Age", "age")
data <- rename.vars(data, "Ethnicity", "ethnicity")
data <- rename.vars(data, "Visit", "visit")
data <- rename.vars(data, "ADOS1", "ados1")
data <- rename.vars(data, "MullenRaw1", "nonverbalIQ")
data <- rename.vars(data, "ExpressiveLangRaw1", "verbalIQ")

data <- data[-c (1, 68, 69, 130, 131, 132, 276, 277, 284, 285), ]
data <- na.omit(data)
```

## Welcome to the fourth exciting part of the Language Development in ASD exercise

In this exercise we will assess how many participants we would need to adequately replicate our findings (ensuring our sample size is adequate, our alpha at 0.05 and our beta at 0.8).

### Exercise 1

How much power does your study have (if your model estimates are quite right)?

- [GitHub]Load your dataset, fit your favorite model, assess power for your main effects and interactions of interest.

```{r + Model}

#Best model from the previous assignments
bestmodel <- lmer(CHI_MLU ~ diagnosis + visit + I(visit^2) + verbalIQ + MOT_MLU + (1 + visit + I(visit^2) | ID), data = data)
summary(bestmodel)


#Changing to interaction model
ritter <- lmer(CHI_MLU ~ diagnosis*(visit + I(visit^2)) + (visit + I(visit^2) | ID), data = data, REML = F)
summary(ritter)

#Plotting the intercation - to see why visit^2 = not significant
fe <- summary(ritter)$coefficients[1:6,1]
fe

plot_df <- expand.grid(visit=c(1,2,3,4,5,6),diagnosis=c(0,1)) %>% 
  mutate(CHI_MLU = fe[1] + fe[2]*diagnosis + fe[3]*visit + fe[4]*visit*visit + fe[5]*diagnosis*visit + fe[6]*diagnosis*visit*visit)

plot_interaction <- ggplot(plot_df, aes(y=CHI_MLU, x = visit, color = as.factor(diagnosis))) + 
  geom_line(size=2) + 
  labs(color = "diagnosis") + 
    scale_color_brewer(palette = "Dark2") + 
  theme(legend.position = "top") + 
  theme_classic()

plot_interaction
#seems very close to linear, esp. ASD



#Linear model
linedanser <- lmer(CHI_MLU ~ diagnosis*visit + verbalIQ + (visit | ID), data = data, REML = F)
summary(linedanser)

```

NOTE: Now with interaction and thereby including diagnosis

```{r + Power calculation}

#Power simulation for diagnosis
#powerD <- powerSim(linedanser, simr::fixed('diagnosisTD', method = 't'), nsim = 200)

#Power simulation for visit
#powerv <- powerSim(linedanser, simr::fixed('visit', method = 't'), nsim = 200)

#Power simulation for visit^2
#powerv2 <- powerSim(bestmodel, simr::fixed('I(visit^2)', method = 't'), nsim = 200)

#Power simulation for interaction
powerVD <- powerSim(linedanser, simr::fixed('diagnosisTD:visit', method = 't'), nsim = 200)

#Power simulation for verbalIQ
powerIQ <- powerSim(linedanser, simr::fixed('verbalIQ', method = 't'), nsim = 200)


#powerD
#powerv
#powerv2

powerVD
powerIQ

lastResult()$warnings

```

- Report the power analysis and comment on what you can (or cannot) use its estimates for.

For this exercise we chose our 'bestmodel' (a bit revised from previous assignments). We chose a quadratic linear mixed effects model predicting child MLU development from number of visit, diagnosis, mother's mean length of utternace and verbalIQ. Moreover, we included visit and diagnosis as an interaction.The model also takes ID as random intercept and random slopes by visit (pseudo-code: Child mean lenght of utterance ~ visit * diagnosis + visit^2 + MOT_MLU + verbalIQ + (1 + visit + visit^2 | ID). 

Power simulations were run on the relevant fixed effects and interaction effect of diagnosis and visits. Assuming that the model has the right estimates, the model's power analysis has the following result:


For the main effects: visit, diagnosis and verbalIQ

Visit
97  %     (93.58, 98.89)
0.10

Diagnosis
86  %     (80.41, 90.49)
-0.42

Verbal IQ
100 %     (98.17, 100.0)
0.075

Interaction (Diagnosis x Visit)
100 %     (98.17, 100.0)
0.25


The power analysis indicates that the study was somewhat either overpowered in the sense that the predictors have a power of nearly 100% which is beyond the threshold. 
Nevertheless, this conclusion cannot be drawn solely based on this power analysis, since it is based on the estimates that our model predicted. These might be overestimates of the true effect size. 

The estimates are interesting in several ways. They can be used a priori to uncover how many participants are needed to ensure meaningful findings. Or - like here - they can be used afterwards to investigate the effect size possible with the number of participants the study is based on.


### Exercise 2

How would you perform a more conservative power analysis?

- Identify and justify a minimum effect size for each of your relevant effects

- [GitHub] take the model from exercise 1 and replace the effects with the minimum effect size that you'd accept.

```{r}
#estimates = fixed effects

#Irrelevant due to significant interaction effect 
fixef(linedanser)["diagnosisTD:visit"]<- 0.15


#Power simulations of set effect sizes
powerVDset <- powerSim(linedanser, simr::fixed('diagnosisTD:visit', method = 't'), nsim = 200)
powerVDset


```


Interaction (Diagnosis x Visit)
98.5 %    (95.68, 99.69)
0.15



- [GitHub] assess the power curve by Child.ID, identifying an ideal number of participants to estimate each effect

```{r}
# assessing power curve for relevant effects

#VerbalIQ
pcIQ <-powerCurve(linedanser, simr::fixed('verbalIQ', method='t'), along = 'ID', nsim = 200)
pcIQ
plot(pcIQ)


#Interaction
pcIA <-powerCurve(linedanser, simr::fixed('diagnosisTD:visit', method='t'), along = 'ID', nsim = 200)
pcIA
plot(pcIA)

#Interaction (finding 80%)
pcIAbreak <-powerCurve(linedanser, simr::fixed('diagnosisTD:visit', method='t'), along = 'ID', nsim = 200, breaks = 25:40)
pcIAbreak
plot(pcIAbreak)

#Visit
#pcV <-powerCurve(linedanser, simr::fixed('visit', method='t'), along = 'ID', nsim = 50)
#pcV
#plot(pcV)

#pc5 <-powerCurve(bestmodel, simr::fixed('I(visit^2)', method='t'), along = 'ID', nsim = 200)
#pc5
#plot(pc5)

#Diagnosis
#pcD <-powerCurve(linedanser, simr::fixed('diagnosisTD', method='t'), along = 'ID', nsim = 50)
#plot(pcD)

```

- [GitHub] if your power estimates do not reach an acceptable threshold simulate additional participants and repeat the previous analysis

- Report the power analysis and comment on what you can (or cannot) use its estimates for.

The estimate of the interaction was lowered in an attempt to make a more conservative power analysis. As our hypotheses refer to the effect of time and diagnosis on child MLU, the interaction effect represents our two main effects and thus remains of primary interest. We consider verbal IQ to be sufficiently conservative as is and have decided to keep it at 0.075. 

Using the chosen minimum effect sizes, these power curves (figures 2-4) indicate that 35 participants would be adequate to reach a threshold of 0.80 for the interaction effect. However, one might notice that if the had study been centred around the effect of verbal IQ, one could argue that ≈ 9 participants would be sufficient


### Exercise 3

Assume you have only the resources to collect 30 kids (15 with ASD and 15 TDs). Identify the power for each relevant effect and discuss whether it's worth to run the study and why

```{r}
set.seed(31)

#list with ASD kids
ASD <- data %>% group_by(ID) %>%
  filter(diagnosis == "ASD", n() == 6)
  
#list with TD kids       
TD <- data %>% group_by(ID) %>%
  filter(diagnosis == "TD", n() == 6)


#15 random IDs in asd-list
asd15 <- sample(unique(ASD$ID), 15)
asd15 <- as.vector(asd15)

#15 random IDs in td-list
td15 <- sample(unique(TD$ID), 15)
td15 <- as.vector(td15)

#list of the 30 rndom selected
all30 <- c(asd15, td15)

#make dataframe with the chosen kids
data30 <- data %>%
  subset(ID %in% all30)


set.seed(44)
#Power analysis
model30 <- lmer(CHI_MLU ~ visit*diagnosis + verbalIQ + (1 + visit | ID), data30, REML = F)

summary(model30)

```


```{r}


#Power simulation for verbalIQ
powerIQ30 <- powerSim(model30, simr::fixed('verbalIQ', method = 't'), nsim = 200)


#Power simulation for interaction
powerVD30 <- powerSim(model30, simr::fixed('visit:diagnosisTD', method = 't'), nsim = 200)


#powerv230 <- powerSim(model30, simr::fixed('I(visit^2)', method = 't'), nsim = 100)
#powerD30 <- powerSim(model30, simr::fixed('diagnosisTD', method = 't'), nsim = 10)


powerIQ30
powerVD30


#Using the minimum effect sizes from before
fixef(model30)["visit:diagnosisTD"]<- 0.15


#accessing power curves
pcIQ30 <-powerCurve(model30, simr::fixed('verbalIQ', method='t'), along = 'ID', nsim = 200)
pcIQ30
plot(pcIQ30)

pcIA30 <-powerCurve(model30, simr::fixed('visit:diagnosisTD', method='t'), along = 'ID', nsim = 200)
pcIA30
plot(pcIA30)

```


Using a subset of 30 random children, 15 ASD and 15 TD and the same model structure as previously, we achieve the following power levels and effect sizes:


Verbal IQ

100 %     (98.17, 100.0)
0.075

Interaction (Diagnosis x Visit)

99.5 %    (97.25, 99.99)
0.22

Using the minimum effect sizes previously selected (changing the interaction to 0.15), our study has the following power levels:

Verbal IQ
100 %     (98.17, 100.0)
0.075

Interaction (Diagnosis x Visit)
86 %      (80.41, 90.49)
0.15

Thus the study remains worth running if one is using a power threshold of 80%. One could argue that these results suggest that the actual study is quite overpowered and could be conducted using far less resources while still capturing the effects.





